---
title: "Exploratory Data Analysis"
output: github_document
date: "2024-10-03"
---

```{r setup, include=FALSE}
library(tidyverse)
```

All of the looking at the data, poking at it between importing and analysis 
Includes: visualisation, checks hypothesis generating (sometimes), mostly about getting a sense of the data before getting started 

Group level summaries today: 
May want to do examine groups visually and quantitatively (can do numeric summaries today)
  Computing a mean, median, sd, missingness in a group 
  
Unless makes the group explicit it doesn't exist especially after you start mutating it etc. group_by makes the grouping that we care about in the dataset explicit so that when we use other things in tidyverse they are aware of it. 

```{r}
weather_df = 
  rnoaa::meteo_pull_monitors(
    c("USW00094728", "USW00022534", "USS0023B17S"),
    var = c("PRCP", "TMIN", "TMAX"), 
    date_min = "2021-01-01",
    date_max = "2022-12-31") |>
  mutate(
    name = case_match(
      id, 
      "USW00094728" ~ "CentralPark_NY", 
      "USW00022534" ~ "Molokai_HI",
      "USS0023B17S" ~ "Waterhole_WA"),
    tmin = tmin / 10,
    tmax = tmax / 10,
    month = lubridate::floor_date(date, unit = "month")) |>
  select(name, id, everything())
```

lubridate to use data and time variables (floor_date can round down, here to the first of the month)

Let's make some plots

```{r}
weather_df %>% 
  ggplot(aes(x = prcp)) + 
  geom_histogram()
```

```{r}
weather_df %>% 
  filter(prcp > 1000)
```

Taking a look at stuff and trying to understand what is going on 

```{r}
weather_df %>%
  filter(tmax > 20, tmax < 30) %>% 
  ggplot(aes(x = tmin, y = tmax, color = name, shape = name)) +
  geom_point()
```

something is weird since Central park and Molokai are recording data differently since their data points are appearing in bands where as Waterhole are more scattered 

##group_by()

```{r}
weather_df %>% 
  group_by(name)
```

```{r}
weather_df %>% 
  group_by(month)
```

Not modifying the dataset, but adding a layer ontop for useful subsequent things 

```{r}
weather_df %>% 
  group_by(name) %>% 
  summarise(n_obs = n())
```

```{r}
weather_df %>% 
  group_by(month) %>% 
  summarise(n_obs = n())
```

the number of distinct observations 

```{r}
weather_df %>% 
  group_by(month) %>% 
  summarise(
    n_obs = n(),
    n_dist = n_distinct(month))
```

```{r}
weather_df %>% 
  group_by(name, month) %>% 
  summarise(
    n_obs = n())
```

One off function that does counting 

```{r}
weather_df %>% 
  count(name)
```

##2x2 tables (kinda)

Binary predictor for something to look like a 2x2 table

```{r}
weather_df %>% 
  drop_na(tmax) %>% 
  filter(name!= "Molokai_HI") %>% 
  mutate(
    cold = case_when(
      tmax < 5 ~ "cold", 
      tmax >= 5 ~ "not_cold"
    )
  ) %>% 
  janitor::tabyl(name, cold)
```

##general numeric summaries 


