---
title: "Vis_2"
output: github_document
date: "2024-10-01"
---

```{r setup, include=FALSE}
library(tidyverse)
library(patchwork)
```

Better graphics today 

```{r}
weather_df = 
  rnoaa::meteo_pull_monitors(
    c("USW00094728", "USW00022534", "USS0023B17S"),
    var = c("PRCP", "TMIN", "TMAX"), 
    date_min = "2021-01-01",
    date_max = "2022-12-31") |>
  mutate(
    name = case_match(
      id, 
      "USW00094728" ~ "CentralPark_NY", 
      "USW00022534" ~ "Molokai_HI",
      "USS0023B17S" ~ "Waterhole_WA"),
    tmin = tmin / 10,
    tmax = tmax / 10) |>
  select(name, id, everything())
```

Make a scatterplot but fancy this time

```{r}
weather_df %>% 
  ggplot(aes(x = tmin, y = tmax, color = name)) + 
  geom_point(alpha = .3) +
  labs(
    title = "Temperature scatterplot",
    x = "Minimum Temp (C)",
    y = "Maximum Temp (C)",
    color = "location",
    caption = "Weather taken from rnoaa package for three stations"
  )
```

Sometimes it repeating information that is in the text already, but that is fine 

Looking at scales -- start with `x` and `y` 

Some things happen by default (the scale of the axis currently from -20 to 20 but we can control that)

```{r}
weather_df %>% 
  ggplot(aes(x = tmin, y = tmax, color = name)) + 
  geom_point(alpha = .3) +
  labs(
    title = "Temperature scatterplot",
    x = "Minimum Temp (C)",
    y = "Maximum Temp (C)",
    color = "location",
    caption = "Weather taken from rnoaa package for three stations"
  ) + 
  scale_x_continuous(
    breaks = c(-15, 0, 20),
    labels = c("-15C", "0", "20")
  ) +
  scale_y_continuous(
    limits = c(0, 30),
    transform = "sqrt"
  )
```

Different ways to get to the same point, using the limits here are a filtering step to only show data from 0 to 30 

Transformations are really helpful (sometimes do not want to directly mutate the dataframe but can transform when plotting the data)

Look at color:

This is not very helpful trying to create your own colour pallete. 
Trying to create your own color scheme usually doesn’t go well

```{r}
weather_df %>% 
  ggplot(aes(x = tmin, y = tmax, color = name)) + 
  geom_point(alpha = .3) +
  labs(
    title = "Temperature scatterplot",
    x = "Minimum Temp (C)",
    y = "Maximum Temp (C)",
    color = "location",
    caption = "Weather taken from rnoaa package for three stations"
  ) + 
  scale_color_hue(h = c(10,40))
```

I encourage you to use the `viridis` package instead. There are several options, but the default color scheme works nicely!

```{r}
weather_df %>% 
  ggplot(aes(x = tmin, y = tmax, color = name)) + 
  geom_point(alpha = .3) +
  labs(
    title = "Temperature scatterplot",
    x = "Minimum Temp (C)",
    y = "Maximum Temp (C)",
    color = "location",
    caption = "Weather taken from rnoaa package for three stations"
  ) + 
  viridis::scale_color_viridis(discrete = TRUE)
```

