---
title: "Vis_2"
output: github_document
date: "2024-10-01"
---

```{r setup, include=FALSE}
library(tidyverse)
library(patchwork)
```

Better graphics today 

```{r}
weather_df = 
  rnoaa::meteo_pull_monitors(
    c("USW00094728", "USW00022534", "USS0023B17S"),
    var = c("PRCP", "TMIN", "TMAX"), 
    date_min = "2021-01-01",
    date_max = "2022-12-31") |>
  mutate(
    name = case_match(
      id, 
      "USW00094728" ~ "CentralPark_NY", 
      "USW00022534" ~ "Molokai_HI",
      "USS0023B17S" ~ "Waterhole_WA"),
    tmin = tmin / 10,
    tmax = tmax / 10) |>
  select(name, id, everything())
```

Make a scatterplot but fancy this time

```{r}
weather_df %>% 
  ggplot(aes(x = tmin, y = tmax, color = name)) + 
  geom_point(alpha = .3) +
  labs(
    title = "Temperature scatterplot",
    x = "Minimum Temp (C)",
    y = "Maximum Temp (C)",
    color = "location",
    caption = "Weather taken from rnoaa package for three stations"
  )
```

Sometimes it repeating information that is in the text already, but that is fine 

Looking at scales -- start with `x` and `y` 

Some things happen by default (the scale of the axis currently from -20 to 20 but we can control that)

```{r}
weather_df %>% 
  ggplot(aes(x = tmin, y = tmax, color = name)) + 
  geom_point(alpha = .3) +
  labs(
    title = "Temperature scatterplot",
    x = "Minimum Temp (C)",
    y = "Maximum Temp (C)",
    color = "location",
    caption = "Weather taken from rnoaa package for three stations"
  ) + 
  scale_x_continuous(
    breaks = c(-15, 0, 20),
    labels = c("-15C", "0", "20")
  ) +
  scale_y_continuous(
    limits = c(0, 30),
    transform = "sqrt"
  )
```

Different ways to get to the same point, using the limits here are a filtering step to only show data from 0 to 30 

Transformations are really helpful (sometimes do not want to directly mutate the dataframe but can transform when plotting the data)

Look at color:

This is not very helpful trying to create your own colour pallete. 
Trying to create your own color scheme usually doesn’t go well

```{r}
weather_df %>% 
  ggplot(aes(x = tmin, y = tmax, color = name)) + 
  geom_point(alpha = .3) +
  labs(
    title = "Temperature scatterplot",
    x = "Minimum Temp (C)",
    y = "Maximum Temp (C)",
    color = "location",
    caption = "Weather taken from rnoaa package for three stations"
  ) + 
  scale_color_hue(h = c(10,40))
```

I encourage you to use the `viridis` package instead. There are several options, but the default color scheme works nicely!

```{r}
weather_df %>% 
  ggplot(aes(x = tmin, y = tmax, color = name)) + 
  geom_point(alpha = .3) +
  labs(
    title = "Temperature scatterplot",
    x = "Minimum Temp (C)",
    y = "Maximum Temp (C)",
    color = "location",
    caption = "Weather taken from rnoaa package for three stations"
  ) + 
  viridis::scale_color_viridis(discrete = TRUE)
```

##Themes 

Before just about how to change what the plot looks in general 
Themes are used to modify non-data elements of a plot – they don’t change mappings or how data are render, but control things like background color and location of the the legend. Using themes can help with general plot appearance.

```{r}
ggp_scatterplot = 
  weather_df %>% 
  ggplot(aes(x = tmin, y = tmax, color = name)) + 
  geom_point(alpha = .3) +
  labs(
    title = "Temperature scatterplot",
    x = "Minimum Temp (C)",
    y = "Maximum Temp (C)",
    color = "location",
    caption = "Weather taken from rnoaa package for three stations"
  ) + 
  viridis::scale_color_viridis(discrete = TRUE)
```

By saying that ggp_scatterplot = something, then can just do ggp_scatterplot later and add things directly onto it 

```{r}
ggp_scatterplot + 
  theme(legend.position = "bottom")
```

Black and white background
```{r}
ggp_scatterplot + 
  theme_bw() + 
  theme(legend.position = "bottom")
```

Order matters here:
Putting in opposite order then legend is on the side, since the theme of black and white will set it back to default 

```{r}
ggp_scatterplot + 
  theme_classic() + 
  theme(legend.position = "bottom")
```

```{r}
ggp_scatterplot + 
  ggthemes::theme_excel() + 
  theme(legend.position = "bottom")
```

**Learning Assessment: Revisit the plot showing tmax against date for each location. Use labels, scale options, and theme changes to improve the readability of this plot**

```{r}
weather_df %>% 
  ggplot(aes(x = date, y = tmax, color = name)) + 
  geom_point() +
  geom_smooth(se = FALSE) +
  viridis::scale_color_viridis(discrete = TRUE) + 
    labs(
      x = "Date",
      y = "Maximum Temp (C)", 
      title = "Seasonal variation in Max Temp",
      size = "Precipitation"
    ) +
    theme_minimal() +
    theme(legend.position = "bottom")
```

General steps: 
1. load data into ggplot
2. Geom_smooth 
3. Color
4. Labels 
5. Theme of the overall 

##Extra bonus stuff in `ggplot`

Use different datasets in different `geom`s

```{r}

```

