---
title: "vis_1"
output: github_document
date: "2024-09-26"
---

```{r setup, include=FALSE}
library(tidyverse)
library(ggridges)
```

rnoaa package (keep track of all the weather data, interact with the data that they make that data public)
Which weather stations that i want to use, which variables, and from which dates

Mutate step: creating new variable called name using case match, re-naming the weather stations 
tmax are to change since before it was in 10ths of a degree change 

```{r}
weather_df = 
  rnoaa::meteo_pull_monitors(
    c("USW00094728", "USW00022534", "USS0023B17S"),
    var = c("PRCP", "TMIN", "TMAX"), 
    date_min = "2021-01-01",
    date_max = "2022-12-31") |>
  mutate(
    name = case_match(
      id, 
      "USW00094728" ~ "CentralPark_NY", 
      "USW00022534" ~ "Molokai_HI",
      "USS0023B17S" ~ "Waterhole_WA"),
    tmin = tmin / 10,
    tmax = tmax / 10) |>
  select(name, id, everything())
```

Making our first plot 
- Define the data set (first)
- Define aesthetic mapping (x and y axis)
- then geometrics ontop 
    Geom_point(put points for all of the data points, scatter plot)

```{r}
ggplot(weather_df, aes(x = tmin, y = tmax)) + 
  geom_point()
```

can also write like this since sometimes want to establish the data frame first then do mutate columns before plotting
```{r}
weather_df %>% 
  ggplot(aes(x = tmin, y = tmax)) +
  geom_point()
```

```{r}
ggp_weather_scatterplot = 
  weather_df %>% 
  ggplot(aes(x = tmin, y = tmax)) + 
  geom_point()

ggp_weather_scatterplot
```

Asking which rows are missing values in their tmax column (Check why some rows are missing)
```{r}
weather_df %>% 
  filter(is.na(tmax))
```

##Fancier scatterplots 

colour is like x and y is a feature of the plot (need to name)
alpha (makes more transparent to see better)
geom_smooth smooth line fitting through the points (SE is standard error)
```{r}
weather_df %>% 
  ggplot(aes(x = tmin, y = tmax, color = name)) + 
  geom_point(alpha = .3, size = .8) + 
  geom_smooth(se = FALSE)
```

Where you define the aesthetics matters
```{r}
weather_df %>% 
  ggplot(aes(x = tmin, y = tmax)) + 
  geom_point(aes(color = name), alpha = .3, size = .8) + 
  geom_smooth(se = FALSE)
```

The first line defines everything that comes out of the data frame (colour = name is applied to everything that comes after, geom_point and smooth)
But geom_smooth does not know anything about colour therefore there is only one smooth line 
Mostly issue is aesthetic mismatch 

Use faceting real quick 

```{r}
weather_df %>% 
  ggplot(aes(x = tmin, y = tmax, color = name)) + 
  geom_point(alpha = .3) + 
  geom_smooth(se = FALSE) +
  facet_grid(. ~ name)
```

dot means that we have don't have anythign seperating by row 
but there is seperating by coloum (selected to seperate by the name column)

Let's make a somewhat more interesting scatterplot
Seeing temperature trends (since date is the x axis) and the size of the dots are representative of precipitation
```{r}
weather_df %>% 
  ggplot(aes(x = date, y = tmax, colour = name, size = prcp)) +
  geom_point(alpha = .3) + 
  geom_smooth(se = FALSE) + 
  facet_grid(. ~ name)
```

##Learning Assessment

focuses only on Central Park, converts temperatures to Fahrenheit, makes a scatterplot of min vs. max temperature, and overlays a linear regression line
```{r}
weather_df %>% 
  filter(name == "CentralPark_NY") %>% 
  mutate(
    tmax_fahr = tmax * (9 / 5) + 32,
    tmin_far = tmin * (9 / 5) + 32
  ) %>% 
  ggplot(aes(x = tmin_far, y = tmax_fahr)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE)
```

The method = lm in geom_smooth give straight line through data points rather than curvy
Data processing piece useful before the visualisation 
