---
title: "Vis_2"
output: github_document
date: "2024-10-01"
---

```{r setup, include=FALSE}
library(tidyverse)
library(patchwork)
library(haven)
```

Better graphics today 

```{r}
weather_df = 
  rnoaa::meteo_pull_monitors(
    c("USW00094728", "USW00022534", "USS0023B17S"),
    var = c("PRCP", "TMIN", "TMAX"), 
    date_min = "2021-01-01",
    date_max = "2022-12-31") |>
  mutate(
    name = case_match(
      id, 
      "USW00094728" ~ "CentralPark_NY", 
      "USW00022534" ~ "Molokai_HI",
      "USS0023B17S" ~ "Waterhole_WA"),
    tmin = tmin / 10,
    tmax = tmax / 10) |>
  select(name, id, everything())
```

Make a scatterplot but fancy this time

```{r}
weather_df %>% 
  ggplot(aes(x = tmin, y = tmax, color = name)) + 
  geom_point(alpha = .3) +
  labs(
    title = "Temperature scatterplot",
    x = "Minimum Temp (C)",
    y = "Maximum Temp (C)",
    color = "location",
    caption = "Weather taken from rnoaa package for three stations"
  )
```

Sometimes it repeating information that is in the text already, but that is fine 

Looking at scales -- start with `x` and `y` 

Some things happen by default (the scale of the axis currently from -20 to 20 but we can control that)

```{r}
weather_df %>% 
  ggplot(aes(x = tmin, y = tmax, color = name)) + 
  geom_point(alpha = .3) +
  labs(
    title = "Temperature scatterplot",
    x = "Minimum Temp (C)",
    y = "Maximum Temp (C)",
    color = "location",
    caption = "Weather taken from rnoaa package for three stations"
  ) + 
  scale_x_continuous(
    breaks = c(-15, 0, 20),
    labels = c("-15C", "0", "20")
  ) +
  scale_y_continuous(
    limits = c(0, 30),
    transform = "sqrt"
  )
```

Different ways to get to the same point, using the limits here are a filtering step to only show data from 0 to 30 

Transformations are really helpful (sometimes do not want to directly mutate the dataframe but can transform when plotting the data)

Look at color:

This is not very helpful trying to create your own colour pallete. 
Trying to create your own color scheme usually doesn’t go well

```{r}
weather_df %>% 
  ggplot(aes(x = tmin, y = tmax, color = name)) + 
  geom_point(alpha = .3) +
  labs(
    title = "Temperature scatterplot",
    x = "Minimum Temp (C)",
    y = "Maximum Temp (C)",
    color = "location",
    caption = "Weather taken from rnoaa package for three stations"
  ) + 
  scale_color_hue(h = c(10,40))
```

I encourage you to use the `viridis` package instead. There are several options, but the default color scheme works nicely!

```{r}
weather_df %>% 
  ggplot(aes(x = tmin, y = tmax, color = name)) + 
  geom_point(alpha = .3) +
  labs(
    title = "Temperature scatterplot",
    x = "Minimum Temp (C)",
    y = "Maximum Temp (C)",
    color = "location",
    caption = "Weather taken from rnoaa package for three stations"
  ) + 
  viridis::scale_color_viridis(discrete = TRUE)
```

##Themes 

Before just about how to change what the plot looks in general 
Themes are used to modify non-data elements of a plot – they don’t change mappings or how data are render, but control things like background color and location of the the legend. Using themes can help with general plot appearance.

```{r}
ggp_scatterplot = 
  weather_df %>% 
  ggplot(aes(x = tmin, y = tmax, color = name)) + 
  geom_point(alpha = .3) +
  labs(
    title = "Temperature scatterplot",
    x = "Minimum Temp (C)",
    y = "Maximum Temp (C)",
    color = "location",
    caption = "Weather taken from rnoaa package for three stations"
  ) + 
  viridis::scale_color_viridis(discrete = TRUE)
```

By saying that ggp_scatterplot = something, then can just do ggp_scatterplot later and add things directly onto it 

```{r}
ggp_scatterplot + 
  theme(legend.position = "bottom")
```

Black and white background
```{r}
ggp_scatterplot + 
  theme_bw() + 
  theme(legend.position = "bottom")
```

Order matters here:
Putting in opposite order then legend is on the side, since the theme of black and white will set it back to default 

```{r}
ggp_scatterplot + 
  theme_classic() + 
  theme(legend.position = "bottom")
```

```{r}
ggp_scatterplot + 
  ggthemes::theme_excel() + 
  theme(legend.position = "bottom")
```

**Learning Assessment: Revisit the plot showing tmax against date for each location. Use labels, scale options, and theme changes to improve the readability of this plot**

```{r}
weather_df %>% 
  ggplot(aes(x = date, y = tmax, color = name)) + 
  geom_point() +
  geom_smooth(se = FALSE) +
  viridis::scale_color_viridis(discrete = TRUE) + 
    labs(
      x = "Date",
      y = "Maximum Temp (C)", 
      title = "Seasonal variation in Max Temp",
      size = "Precipitation"
    ) +
    theme_minimal() +
    theme(legend.position = "bottom")
```

General steps: 
1. load data into ggplot
2. Geom_smooth 
3. Color
4. Labels 
5. Theme of the overall 

##Extra bonus stuff in `ggplot`

Use different datasets in different `geom`s

```{r}
central_park_df = 
  weather_df |> 
  filter(name == "CentralPark_NY")

molokai_df = 
  weather_df |> 
  filter(name == "Molokai_HI")

ggplot(data = molokai_df, aes(x = date, y = tmax, color = name)) + 
  geom_point() + 
  geom_line(data = central_park_df) 
```

Overlaying two different geometries, where one is a line and another is a dot can overlay two differnt geom types 

##Multiple panels 

```{r}
weather_df %>% 
  ggplot(aes(x = tmax, fill = name)) + 
  geom_density() + 
  facet_grid(. ~ name)
```

All the plot type (density plots) but in seperate facet grids (here seperated out by name)

```{r}
ggp_tmax_tmin = 
  weather_df %>% 
  ggplot(aes(x = tmin, y = tmax, color = name)) +
  geom_point(alpha = .3)

ggp_tmax_density = 
  weather_df %>% 
  ggplot(aes(x = tmax, fill = name)) + 
  geom_density(alpha = .3)

ggp_tmax_date = 
  weather_df %>% 
  ggplot(aes(x = date, y = tmax, color = name)) +
  geom_point() +
  geom_smooth(se = FALSE)

(ggp_tmax_date + ggp_tmax_density) / ggp_tmax_date
```

Here there are two different types of graphs, one is a scatter and another is a density function, cannot overlay 
Handy to say that here is one type of information here is another type of information 

##Data Manipulation 

What if we want to assign different colors to be different things?

```{r}
weather_df %>% 
  ggplot(aes(x = name, y = tmax, fill = name)) + 
  geom_violin(alpha = .5)
```

Question: how does it know what order to put onto the x-axis 

When take a categorical variable, and make it onto the x-axis, R needs to know what order to plot them (1st, 2nd, 3rd)
By default is alphabetical order

Work with the factor variables that happens in the background of R that helps decide these things, if we can intervene with these factor variables then we can manipulate 

```{r}
weather_df %>% 
  mutate(name = fct_relevel(name, c("Molokai_HI", "CentralParkNY", "Waterhole_WA"))) %>% 
  ggplot(aes(x = name, y = tmax, fill = name)) + 
  geom_violin(alpha = .5)
```

Very explicit that it is not in alphabetical order anymore but in this one 

One of the thing that ggplot is choosing is the factor variable and changing the factor variable can help better manipulate 


##PULSE data next 

Imagine a collection of box plots showing the distribution of BDI score at 1,6,12 months

```{r}
pulse_df = 
  read_sas("data/public_pulse_data.sas7bdat") %>% 
  janitor::clean_names()
```

We cannot do this with this dataset as it is right now since there is missing data, I don't have a visit variable. Since the pulse_df is that each visit is a seperate variable (bdi_score_01m, bdi_score_06m and bdi_score_12m). 
Need to pivot longer

```{r}
pulse_df = 
  read_sas("data/public_pulse_data.sas7bdat") %>% 
  janitor::clean_names() %>% 
  pivot_longer(
    cols = bdi_score_bl:bdi_score_12m,
    names_to = "visit",
    values_to = "bdi_score",
    names_prefix = "bdi_score_"
  ) %>% 
  mutate(visit = ifelse(visit == "bl", "00m", visit))

pulse_df %>% 
  ggplot(aes(x = visit, y = bdi_score)) + 
  geom_boxplot()
```

take each bdi, stack them up, new column called visit and have new rows for baseline, 1m, 6m and 12m as new inputs. 

**Learning Assessment**

```{r}
litters_df = 
  read_csv("data/FAS_litters.csv", na = c("NA", ".", "")) %>% 
  janitor::clean_names() %>% 
  separate(group, into = c("dose", "tx_day"), 3)

pups_df = 
  read_csv("data/FAS_pups.csv", na = c("NA", ".", "")) %>% 
  janitor::clean_names() %>% 
  pivot_longer(
    pd_ears:pd_walk,
    names_to = "outcome",
    values_to = "pn_day",
    names_prefix = "pd_"
  )

fas_df = 
  left_join(pup_data, litter_data, by = "litter_number") 
```


